Historic collections as a tool for assessing the global pollination crisis

### 2018-04-06 ms.Rmd
Last touches 2

*Affiliations:*
^1^ Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, E-41092 Sevilla, Spain   
^2^ Centre for Biodiversity and Biosecurity, School of Biological Sciences, The University of Auckland, Auckland, New Zealand  
^3^ Landcare Research, Auckland, New Zealand  
^4^ Andrena Iniciativas y Estudios Medio Ambientales
*Correspondence: nacho.bartomeus@gmail.com

@@

One solution to overcoming these barriers is the use of space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach [@Franca2016; @DePalma2018], these studies currently provide the most extensive source of pollinator population data. For example, researchers have recently estimated bee richness declines for every country in Europe using predictions from models of pollinator associations with different land-use types [@DePalma2017]. A second important method is the use of data collected from pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs [@vanSwaay2008] and is currently being extended to other pollinator taxa. However, these programs require significant time to generate long-term datasets and cannot be used to assess historic pollinator populations. Finally, the most robust approach for assessing long-term historical pollinator population trends is to use historical information on species occurrences, which is often archived in museum collections [e.g., @Bartomeus2013a].

@@

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
write.csv(lhf, "data/SpMat2.csv")
#plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
#abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

@@

#table?
#lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) < 7.5),]
#lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),]
#nrow(lhf[which(log(lhf$post1980) < 6.9 & log(lhf$pre1980) < 6.9),])
# poland switcherland need old records! > 10000 vs 300
# sri lanka and french guayana, 
# nicaragua, venezuela, Ecuador..>2000 vs >500.
# Colombia & Panama, zimbawe > 3000 vs >1500
#lhf[which(lhf$country == "ES" | lhf$country == "NZ"),]
col <- ifelse(lhf$country == "ES" | lhf$country == "NZ", "red", "black")
col <- ifelse(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5, col, "grey")
pchs <- ifelse(lhf$country == "ES" | lhf$country == "NZ", 19, 1)
plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col, pch = pchs)
temp <- subset(lhf, country %in% c("ES", "NZ"))
text(x = temp$log_post1980, y = temp$log_pre1980+2 , labels = temp$country, col = "red")
#36 countries!
#exp(7.5) #> 1800 records in both axes!
```

```{r get rgbif, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
#Can we do rar richness per time period $ country and test? YES!! 
#1) Select countries.
countries <- lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),"country"]
#2) Get all data
occ_list <- list()
for(i in 1:length(countries)){
    taxon_list <- list()
    count_list <- list()
    for(j in 1:length(keys)){
        taxon_list[[j]] <- occ_data(taxonKey= keys[j], 
                            year = '1850,2018', 
                            country = as.character(countries[i]),
                            limit = 200000)
        #count_list[[j]] <- occ_count(taxonKey= keys[j],
                         #    from = 1850, to = 1979,
                         #   georeferenced=TRUE,
                         #   country = countries[i])
    }
    select <- c("name", "year")
    out <- rbind(as.data.frame(taxon_list[[1]]$data[,select]), 
                 as.data.frame(taxon_list[[2]]$data[,select]),
                 as.data.frame(taxon_list[[3]]$data[,select]),
                 as.data.frame(taxon_list[[4]]$data[,select]),
                 as.data.frame(taxon_list[[5]]$data[,select]),
                 as.data.frame(taxon_list[[6]]$data[,select]),
                 as.data.frame(taxon_list[[7]]$data[,select]),
                 as.data.frame(taxon_list[[8]]$data[,select]))
    occ_list[[i]] <- out
    print(i)
}
#save output
save(occ_list, file = "data/gbif.RData")

@@

#map
data2$change <- ((data2$post1980 - data2$pre1980)*100 / data2$pre1980)

map.gbif <- joinCountryData2Map(data2, joinCode="ISO2", nameJoinColumn = "countries",
                            nameCountryColumn = "countries")

#average richness change
colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
colorss <- (heat.colors(10))
map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette= colorss,
                       mapTitle = "Average bee richness change (%)", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))

@@

```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
dat <- read.csv("data/spain_clean_used.csv")
#Repeat richness and genus
rich_tot <- dcast(dat, when ~ ., value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#rich_tot #wow.
#extract genus
pos <- regexpr(dat$matched_name2, pattern = " ", fixed = TRUE)
dat$genus <- substr(dat$matched_name2, 
                    start = 1,
                    stop = pos-1)
genus_tot <- dcast(dat, when ~ ., value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
genus_tot2 <- dcast(dat,  genus ~ when, value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot #wow, lot's of genus lost!
#genus_tot2
#subset(genus_tot2, now == 0 & asensio == 1) #20 lost!
#subset(genus_tot2, now == 1 & asensio == 1) #27 stable
#subset(genus_tot2, now == 1 & asensio == 0) #2 added!
genus_tot3 <- dcast(dat,  genus ~ when, value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot3
#colSums(genus_tot3[,2:3])
#write.csv(genus_tot3, "raw_data/spain_species_per_genus.csv")

>>>>

*Affiliations:*
^1^ Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, E-41092 Sevilla, Spain   
^2^ Centre for Biodiversity and Biosecurity, School of Biological Sciences, The University of Auckland, Auckland, New Zealand  
^3^ Landcare Research, Auckland, New Zealand  
^4^ Andrena Iniciativas y Estudios Medio Ambientales, Valladolid, Spain  
*Correspondence: nacho.bartomeus@gmail.com

@@

One solution to overcoming these barriers is the use of space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach [@Franca2016; @DePalma2018], these studies currently provide the most extensive source of pollinator population data. For example, researchers have recently estimated bee richness declines for every country in Europe using predictions from models of pollinator associations with different land-use types [@DePalma2017]. A second important method is the use of data collected from pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs [@vanSwaay2008] and is currently being extended to other pollinator taxa. However, these programs require significant time to generate long-term datasets and cannot be used to assess historic pollinator populations. Finally, the most practical approach for assessing long-term historical pollinator population trends is to use historical information on species occurrences, which is often archived in museum collections [e.g., @Bartomeus2013a].

@@

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
#write.csv(lhf, "data/SpMat2.csv")
#plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
#abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

@@

#table?
#lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) < 7.5),]
#lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),]
#nrow(lhf[which(log(lhf$post1980) < 6.9 & log(lhf$pre1980) < 6.9),])
# poland switcherland need old records! > 10000 vs 300
# sri lanka and french guayana, 
# nicaragua, venezuela, Ecuador..>2000 vs >500.
# Colombia & Panama, zimbawe > 3000 vs >1500
#lhf[which(lhf$country == "ES" | lhf$country == "NZ"),]
col <- ifelse(lhf$country == "ES" | lhf$country == "NZ", "red", "black")
col <- ifelse(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5, col, "grey")
pchs <- ifelse(lhf$country == "ES" | lhf$country == "NZ", 19, 1)
plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col, pch = pchs, las = 1)
temp <- subset(lhf, country %in% c("ES", "NZ"))
text(x = temp$log_post1980, y = temp$log_pre1980+2 , labels = temp$country, col = "red")
abline(h = 7.5, v = 7.5)
#36 countries!
#exp(7.5) #> 1800 records in both axes!
```

```{r get rgbif, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
#Can we do rar richness per time period $ country and test? YES!! 
#1) Select countries.
countries <- lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),"country"]
#2) Get all data
occ_list <- list()
for(i in 1:length(countries)){
    taxon_list <- list()
    count_list <- list()
    for(j in 1:length(keys)){
        taxon_list[[j]] <- occ_data(taxonKey= keys[j], 
                            year = '1850,2018', 
                            country = as.character(countries[i]),
                            limit = 200000)
        #count_list[[j]] <- occ_count(taxonKey= keys[j],
                         #    from = 1850, to = 1979,
                         #   georeferenced=TRUE,
                         #   country = countries[i])
    }
    select <- c("name", "year")
    out <- rbind(as.data.frame(taxon_list[[1]]$data[,select]), 
                 as.data.frame(taxon_list[[2]]$data[,select]),
                 as.data.frame(taxon_list[[3]]$data[,select]),
                 as.data.frame(taxon_list[[4]]$data[,select]),
                 as.data.frame(taxon_list[[5]]$data[,select]),
                 as.data.frame(taxon_list[[6]]$data[,select]),
                 as.data.frame(taxon_list[[7]]$data[,select]),
                 as.data.frame(taxon_list[[8]]$data[,select]))
    occ_list[[i]] <- out
    print(i)
}
#save output
#save(occ_list, file = "data/gbif.RData")

@@

#map
data2$change <- ((data2$post1980 - data2$pre1980)*100 / data2$pre1980)
#mean(data2$change)
map.gbif <- joinCountryData2Map(data2, joinCode="ISO2", nameJoinColumn = "countries",
                            nameCountryColumn = "countries")

#average richness change
#colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
colorss <- (heat.colors(10))
map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette= colorss,
                       mapTitle = "Average bee richness change (%)", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))

@@

```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
dat <- read.csv("data/spain_clean_used.csv")
#Repeat richness and genus
rich_tot <- dcast(dat, when ~ ., value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#rich_tot #wow.
#extract genus
pos <- regexpr(dat$matched_name2, pattern = " ", fixed = TRUE)
dat$genus <- substr(dat$matched_name2, 
                    start = 1,
                    stop = pos-1)
genus_tot <- dcast(dat, when ~ ., value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
genus_tot2 <- dcast(dat,  genus ~ when, value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot #wow, lot's of genus lost!
#genus_tot2
#subset(genus_tot2, now == 0 & asensio == 1) #20 lost!
#subset(genus_tot2, now == 1 & asensio == 1) #27 stable
#subset(genus_tot2, now == 1 & asensio == 0) #2 added!
genus_tot3 <- dcast(dat,  genus ~ when, value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot3
#colSums(genus_tot3[,2:3])
#write.csv(genus_tot3, "data/SupMat4.csv")
